### 2022-06-19 chapter7/chapter.tex
Ch7 2nd draft

%!TEX root = ../thesis.tex
\chapter{Conclusion\label{ch: conclusion}} 
A fluid-fluid queue is a stochastic fluid queue, where the driving process is a fluid queue itself. Fluid queues provide a model for a single continuous performance measure of a system in the presence of a random environment. Fluid queues have found a wide variety of applications including risk processes, telecommunications, and environmental modelling, among others. Given the success of fluid queues it is believed that the extension to fluid-fluid queues, which enable us to track two continuous performance measures of a system, will also find success. \cite{bo2014} provide an analysis of fluid-fluid queues and derive operator-analytic expressions for the first-return operator, and stationary distribution of a fluid-fluid queue.

>>>>

%!TEX root = ../thesis.tex
\chapter{Conclusion\label{ch: conclusion}} 
A fluid-fluid queue is a stochastic fluid queue, where the driving process is a fluid queue itself. Fluid queues provide a model for a single continuous performance measure of a system in the presence of a random environment. Fluid queues have found a wide variety of applications including risk processes, telecommunications, and environmental modelling, among others. Given the success of fluid queues it is believed that the extension to fluid-fluid queues, which enable us to track two continuous performance measures of a system, will also find success. \cite{bo2014} provide an analysis of fluid-fluid queues and derive operator-analytic expressions for the first-return operator, and limiting distribution of a fluid-fluid queue.

### 2022-06-19 front/abstract.tex
Ch7 2nd draft

This thesis provides approximations to fluid queue so that we can approximate the operators in \cite{bo2014}. It investigates three main approximation schemes; the DG scheme (Chapter~\ref{ch:galerkin}) which is a popular finite-element scheme, the uniformisation scheme of \cite{bo2013} which approximates a fluid queue but a continuous-time Markov chain (specifically, a quasi-birth-and-death-process (QBD)), and the QBD-RAP scheme which is a generalisation of a QBD to allow matrix exponential inter event times. The QBD-RAP scheme is novel; we describe the construction of the scheme in Chapter~\ref{sec: construction and modelling}, and provide an analysis to show that it is convergent in Chapters~\ref{sec: conv} and \ref{ch: global results}. We demonstrate the effectiveness of the approximation schemes in Chapter~\ref{sec: numerics}, focussing on problems with discontinuous solutions. In general, we find that the DG scheme performs remarkably well for smooth problems, but can produce oscillations and negative probability estimates in the presence of discontinuities, the QBD-RAP approximation performs well in the presence of discontinuities, but does not perform as well as the DG scheme for smooth problems, the uniformisation scheme produces reliable approximations in the presence of discontinuities, but converges slowest for all problems.

>>>>

This thesis provides approximations to fluid queues so that we can approximate the operators in \cite{bo2014}. It investigates three main approximation schemes; the DG scheme (Chapter~\ref{ch:galerkin}) which is a popular finite-element scheme, the uniformisation scheme of \cite{bo2013} which approximates a fluid queue by a continuous-time Markov chain (specifically, a quasi-birth-and-death-process (QBD)), and the QBD-RAP scheme which is a generalisation of a QBD to allow matrix exponential inter event times. The QBD-RAP scheme is novel; we describe the construction of the scheme in Chapter~\ref{sec: construction and modelling}, and provide an analysis to show that it is convergent in Chapters~\ref{sec: conv} and \ref{ch: global results}. We demonstrate the effectiveness of the approximation schemes in Chapter~\ref{sec: numerics}, focussing on problems with discontinuous solutions. In general, we find that the DG scheme performs remarkably well for smooth problems, but can produce oscillations and negative probability estimates in the presence of discontinuities, the QBD-RAP approximation performs well in the presence of discontinuities, but does not perform as well as the DG scheme for smooth problems, the uniformisation scheme produces reliable approximations in the presence of discontinuities, but converges slowest for all problems.

### 2022-06-19 intro/chapter.tex
Ch7 2nd draft

More recently, \cite{bo2014} extends fluid queues to so-called \emph{stochastic fluid-fluid queues}. In a fluid-fluid queue there is a second level process, \(\{Y(t)\}_{t\geq0}\) which is itself driven by a fluid queue, \(\{(X(t),\varphi(t))\}_{t\geq0}\), so \((X(t),\varphi(t))\) determines the rate at which \(\{Y(t)\}\) moves. The addition of the second level process \(\{Y(t)\}\) greatly increases the modelling potential. For example, \cite{hydro,bo2014} describe a model for a hydroelectric power generator which can be operated `on design' and `off design'. When the system operates `off design' there system is less efficient and there is more wear, but may be used to optimise overall system performance. Further, maintenance of the generator takes place periodically to improve its performance and lifespan. The system can plausibly be modelled as a fluid queue, \(\{(\varphi(t),X(t))\}\), where the state space of \(\{\varphi(t)\}\) consists of states representing the systems states `start', `stop', `on design', `off design', `maintenance', and `idle'. The level variable \(\{X(t)\}\) take values in \([0,1]\) and represents the deterioration level of the system with 0 being perfect working order and 1 meaning the system needs replacement. The second level process \(\{Y(t)\}\) could be used to model the revenue of the power generator, which, logically, should depend on the state of the system, \(\{\varphi(t)\}\). Moreover, the revenue could also depend on the deterioration level \(\{X(t)\}\), as, for example, when the system is brand new it may operate more efficiently, and when it is near broken it may be less efficient.

@@

The addition of the second level process \(\{Y(t)\}\) greatly increases the flexibility and modelling potential, however comes at some cost to the ease of their analysis. The analysis of \cite{bo2014} is in principle similar to the matrix-analytic methods of \cite{bean2005}, and derives results about the second level process \(\{Y(t)\}_{t\geq0}\) in terms of the infinitesimal generator (a differential operator) of the fluid queue, \(\{(X(t),\varphi(t))\}_{t\geq0}\). For practical computation of the results of \cite{bo2014}, a discretisation of the infinitesimal generator of the fluid queue can be used. To this end, to date, two possible discretisation have been suggested. Taking a differential equations-based approach, \cite{blnos2022} use the discontinuous Galerkin (DG) method to discretise the operator, while \cite{bo2013} take a stochastic modelling and matrix-analytic methods approach to approximate the fluid-queue by a quasi-birth-and-death (QBD) process. The QBD approximation of \cite{bo2013} is derived via a uniformisation argument, so we refer to it as the uniformisation scheme throughout this thesis. Both approaches are insightful and offer different tools and perspectives with which to analyse the resulting approximations. It turns out that the uniformisation scheme of \cite{bo2013} is a subclass of the former; the uniformisation scheme can be viewed as the simplest DG scheme where cells are all have equal width, the operator is projected onto a basis of piecewise constant functions, and an upwind flux is used to approximate the flow of mass between cells. This is also equivalent to a finite-volume scheme.

@@

Furthermore, for the DG method there is a limited theoretical backing as to why we may approximate the operator-Riccati equation with the matrix-Riccati equation other than it is a sensible thing to do, and it seems to work in practice. In the case of the QBD approximation of \cite{bo2013} the approximate matrix-Riccati is more justified as in it can be derived by considering the first-return probabilities for a fluid queue driven by the approximating QBD where the rates of the fluid are determined by a piecewise constant approximation to \(r_i(x)\) \citep{bean2005}.

@@

As the QBD-RAP has a stochastic interpretation then the approximations it produces are guaranteed to have non-negative density functions. Further, for the QBD-RAP (and uniformisation) scheme non-negativity is guaranteed for any initial condition without any further computation or post-processing. For the approximation of fluid-fluid queues, the matrix-Riccati equation we solve to approximate the first-return distribution of a fluid-fluid queue can be derived by considering the first-return probabilities for a RAP-modulated fluid queue driven by the QBD-RAP \citep{p2019,bgnp2021}. The stochastic interpretation can also be leveraged to aid in the analysis of the approximation. Another attractive property is that the approximations to operators obtained via the QBD-RAP (and uniformisation) scheme are linear which is necessary for the application to approximating the first-return operator of fluid-fluid queues and may also provide other computational and analytical benefits.

@@

Chapter~\ref{ch: global results} then looks to extend the local convergence results of Chapter~\ref{sec: conv} to a global convergence result on the whole domain of approximation. Chapter~\ref{ch: global results} uses more traditional Markov process arguments which rely on things such as the strong Markov property, time-homogeneity and the Law of total probability. In Chapter~\ref{ch: global results} we first consider the discrete-time process which the process embedded in the QBD-RAP at times when the QBD-RAP sees an orbit restart epoch. We prove that the embedded process converges in distribution to a corresponding embedded process of the fluid queue. The rest of Chapter~\ref{ch: global results} then attempt to prove a global converge result. The main result being Theorem~\ref{thm: big thm} which states that the QBD-RAP approximation scheme converges weakly (weakly with respect to the spatial and temporal variable) to the distribution of the fluid queue.

@@

To help the reader understand the notations used in the DG scheme, we provide a small toy model example in Appendix~\ref{appendix:example}. Appendix~\ref{sec:properties} provides a proof that the DG scheme conserves probability. The rest of the appendices contain technical results relating to proving the convergence of the QBD-RAP approximation scheme. Appendix~\ref{appendix: sec: 2} proves that the \emph{closing operators} introduced as part of the QBD-RAP scheme have the properties we claim they do to prove convergence in Chapter~\ref{sec: conv}. Appendix~\ref{app:extend conv} extends some results from Chapter~\ref{sec: conv} to a setting which requires slightly less computation. Lastly, Appendix~\ref{appendix: kronecker} provides some algebraic results which help us to manipulate certain Laplace transforms from Chapter~\ref{sec: conv}.

>>>>

% Of particular relevance to this thesis is the matrix-analytic analysis of fluid queues in \cite{bean2005}. Sample paths of fluid queues can be broken up into contiguous (random) intervals of time on which either, the phase remains in some subset of phases, \(\varphi(t) \in \calS_+'\) say, and the level process of the fluid queue is non-decreasing, or the phase remains in some subset of phases, \(\varphi(t) \in \calS_-'\) say, and the level process of the fluid queue is non-increasing. For \(m\in\{+,-\}\), the distribution of 
% \begin{align}
% 	&\mathbb P(X(t)\geq x, \varphi(t)=j\in\calS_m', t\leq E^\lambda \mid X(0)=x_0, \varphi(0)=i\in\calS_m') \label{eqn: LLLppL}
% \end{align}
% where \(E^\lambda\) is an exponential random variable with rate \(\lambda\) is readily computable as a Phase-type distribution. In particular, the probability (\ref{eqn: LLLppL}) is characterised in terms of the \emph{generator} of the phase process, and the rates at which the level process moves in each phase. %Moreover, taking the Laplace transform with respect to time of the distribution yields 
% \begin{align*}
% 	\mathbb P(X(t)\geq x, \varphi(u)\in\calS_+, u\in[0,t), \varphi(t)=j\in\calS_+, t\leq E^\lambda \mid \varphi(0)=i\in\calS_+)
% \end{align*}

More recently, \cite{bo2014} extend fluid queues to so-called \emph{stochastic fluid-fluid queues}. In a fluid-fluid queue there is a second level process, \(\{Y(t)\}_{t\geq0}\) which is itself driven by a fluid queue, \(\{(X(t),\varphi(t))\}_{t\geq0}\), so \((X(t),\varphi(t))\) determines the rate at which \(\{Y(t)\}\) moves. The addition of the second level process \(\{Y(t)\}\) greatly increases the modelling potential. For example, \cite{hydro,bo2014} describe a model for a hydroelectric power generator which can be operated `on design' and `off design'. When the system operates `off design' the system is less efficient and there is more wear, but may be used to optimise overall system performance. Further, maintenance of the generator takes place periodically to improve its performance and lifespan. The system can plausibly be modelled as a fluid queue, \(\{(\varphi(t),X(t))\}\), where the state space of \(\{\varphi(t)\}\) consists of states representing the systems states `start', `stop', `on design', `off design', `maintenance', and `idle'. The level variable \(\{X(t)\}\) take values in \([0,1]\) and represents the deterioration level of the system with 0 being perfect working order and 1 meaning the system needs replacement. The second level process \(\{Y(t)\}\) could be used to model the revenue of the power generator, which, logically, should depend on the state of the system, \(\{\varphi(t)\}\). Moreover, the revenue could also depend on the deterioration level \(\{X(t)\}\), as, for example, when the system is brand new it may operate more efficiently, and when it is near broken it may be less efficient.

@@

The addition of the second level process \(\{Y(t)\}\) greatly increases the flexibility and modelling potential. However, the added flexibility comes at a cost to the in the analysis of the model. The analysis of \cite{bo2014} is in principle similar to the matrix-analytic methods of \cite{bean2005}, and derives results about the second level process \(\{Y(t)\}_{t\geq0}\) in terms of the infinitesimal generator (a differential operator) of the fluid queue, \(\{(X(t),\varphi(t))\}_{t\geq0}\). For practical computation of the results of \cite{bo2014}, a discretisation of the infinitesimal generator of the fluid queue can be used. To this end, to date, two possible discretisations have been suggested. Taking a differential equations-based approach, \cite{blnos2022} use the discontinuous Galerkin (DG) method to discretise the operator, while \cite{bo2013} take a stochastic modelling and matrix-analytic methods approach to approximate the fluid-queue by a quasi-birth-and-death (QBD) process. The QBD approximation of \cite{bo2013} is derived via a uniformisation argument, so we refer to it as the uniformisation scheme throughout this thesis. Both approaches are insightful and offer different tools and perspectives with which to analyse the resulting approximations. It turns out that the uniformisation scheme of \cite{bo2013} is a subclass of the former; the uniformisation scheme can be viewed as the simplest DG scheme where cells are all have equal width, the operator is projected onto a basis of piecewise constant functions, and an upwind flux is used to approximate the flow of mass between cells. This is also equivalent to a finite-volume scheme.

@@

Furthermore, for the DG method there is a limited theoretical backing as to why we may approximate the operator-Riccati equation with the matrix-Riccati equation other than it is a sensible thing to do, and it seems to work in practice. In the case of the QBD approximation of \cite{bo2013} the approximate matrix-Riccati equation is more justified as in it can be derived by considering the first-return probabilities for a fluid queue driven by the approximating QBD where the rates of the fluid are determined by a piecewise constant approximation to \(r_i(x)\) \citep{bean2005}.

@@

As the QBD-RAP has a stochastic interpretation the approximations it produces are guaranteed to have non-negative density functions. Further, for the QBD-RAP (and uniformisation) scheme non-negativity is guaranteed for any initial condition without any further computation or post-processing. For the approximation of fluid-fluid queues, the matrix-Riccati equation we solve to approximate the first-return distribution of a fluid-fluid queue can be derived by considering the first-return probabilities for a RAP-modulated fluid queue driven by the QBD-RAP \citep{p2019,bgnp2021}. The stochastic interpretation can also be leveraged to aid in the analysis of the approximation. Another attractive property is that the approximations to operators obtained via the QBD-RAP (and uniformisation) scheme are linear which is necessary for the application to approximating the first-return operator of fluid-fluid queues and may also provide other computational and analytical benefits.

@@

Chapter~\ref{ch: global results} then looks to extend the local convergence results of Chapter~\ref{sec: conv} to a global convergence result on the whole domain of approximation. Chapter~\ref{ch: global results} uses more traditional Markov process arguments which rely on things such as the strong Markov property, time-homogeneity and the Law of total probability. In Chapter~\ref{ch: global results} we first consider the discrete-time process which the process embedded in the QBD-RAP at times when the QBD-RAP sees an orbit restart epoch. We prove that the embedded process converges in distribution to a corresponding embedded process of the fluid queue. The rest of Chapter~\ref{ch: global results} then attempt to prove a global converge result. The main result being Theorem~\ref{thm: big thm} which states that the QBD-RAP approximation scheme converges weakly (with respect to the spatial and temporal variable) to the distribution of the fluid queue.

@@

To help the reader understand the notation used in the DG scheme, we provide a small toy model example in Appendix~\ref{appendix:example}. Appendix~\ref{sec:properties} provides a proof that the DG scheme conserves probability. The rest of the appendices contain technical results relating to proving the convergence of the QBD-RAP approximation scheme. Appendix~\ref{appendix: sec: 2} proves that the \emph{closing operators} introduced as part of the QBD-RAP scheme have the properties we claim they do to prove convergence in Chapter~\ref{sec: conv}. Appendix~\ref{app:extend conv} extends some results from Chapter~\ref{sec: conv} to a setting which requires slightly less computation. Lastly, Appendix~\ref{appendix: kronecker} provides some algebraic results which help us to manipulate certain Laplace transforms from Chapter~\ref{sec: conv}.
